$p$-adic precision

### 2014-02-15 padicprec.tex
Some remarks about non-surjectivity of differentials (to be continued)

\begin{deftn}
\label{def:firstorder}
Let $E$ and $F$ be two $K$-Banach and $f : U \rightarrow F$ be a 
function defined on an open subset $U$ of $E$. Let also $v_0$ be a 
point in $U$.
A lattice $H$ in $E$ is called a \emph{first order lattice} for $f$ at 
$v_0$ if the equality
$$f(v_0 + H) = f(v_0) + f'(v_0) (H)$$
holds.
\end{deftn}

@@

Let us assume now that $f$ is locally analytic on $U$ and $H$ is a first 
order lattice for $f$ at $x$. Then $f(x+H) = f(x) + f'(x)(H)$ and the 
smallest possible $H'$ is $f'(x)(H)$. In other words, the function {\tt 
f} generates artificial loss of precision if the precision $H'$ it 
outputs is strictly larger that $f'(x)(H)$. The aim of this section is 
to explain how, under the above hypothesis, one can implement the 
function {\tt f} so that it always outputs the optimal precision $O(H')$ 
with $H' = f'(x)(H)$.

@@

\subsubsection*{Quick comparison with floating point arithmetics}

It worths remarking that the two strategy designed above share many 
similarities with usual floating point arithmetics over the reals. 
Indeed, roughly speaking, in each setting, we begin by choosing a large 
precision, we do all our computations up to this precision understood 
that when we are not sure about some digit, we choose it ``at random'' 
or using good heuristics. The main difference is that, in the 
ultrametric setting, we are able (under some mild hypothesis) to 
quantify the precision we need at each individual step in order to be 
sure that the final result is correct up to the required precision.

>>>>

\begin{deftn}
\label{def:firstorder}
Let $E$ and $F$ be two $K$-Banach and $f : U \rightarrow F$ be a 
function defined on an open subset $U$ of $E$. Let also $v_0$ be a 
point in $U$.
A lattice $H$ in $E$ is called a \emph{first order lattice} for $f$ at 
$v_0$ if the equality
\begin{equation}
\label{eq:firstorder}
f(v_0 + H) = f(v_0) + f'(v_0) (H)
\end{equation}
holds.
\end{deftn}

We emphasize that we require an equality in Eq.~\eqref{eq:firstorder}, 
and not just an inclusion! It will be very important for the 
applications we have in mind.
Here comes our main Lemma.

@@

Let us assume now that $f$ is locally analytic on $U$ and in addition 
that $f'(x)$ is surjective. Proposition \ref{prop:locanalytic} then 
yields a rather simple sufficient condition to decide if a given lattice 
$H$ is a first order lattice for $f$ at $x$. For such a lattice, by 
definition, we have $f(x+H) = f(x) + f'(x)(H)$. Thus the smallest 
possible $H'$ is $f'(x)(H)$. In other words, the function {\tt f} 
generates artificial loss of precision if the precision $H'$ it outputs 
is strictly larger that $f'(x)(H)$. The aim of this section is to 
explain how, under the above hypothesis, one can implement the function 
{\tt f} so that it always outputs the optimal precision $O(H')$ with $H' 
= f'(x)(H)$.

@@

\subsubsection*{Additional remarks}

From the beggining, we have assumed that $f'(x)$ is surjective. Let us 
discuss shortly what happens when this assumption is relaxed. As it is 
explained after the proof of Lemma \ref{lem:main}, the first thing we 
can do is to project the result onto different quotients, \emph{i.e.} to 
work with the composites $\pr_K \circ f$ for a sufficiently large family 
of closed sub vector spaces $K \subset F$ such that $K + f'(x)(E) = F$. 
(We recall that here $\pr_K$ denotes the canonical projection to $F$ 
onto the quotient $F/K$.) If $F$ has a natural system of coordinates 
(which is the case in almost all settings), we can generally take the 
$\pr_K$'s as the projections on each coordinate. Doing this, we end up 
with a precision on each individual coordinate. Furthermore, we have the 
guarantee that all these precisions are sharp (even if the lattice build
from them is not).
Let us illustrate the above discussion by an example: suppose that we 
want to compute the function $f : (K^n)^n \to M_n(K)$ that takes a 
family of $n$ vectors to its Graam matrix. The differential of $f$ is
clearly never surjective because $f$ takes its values in the subspace
consisting of symmetric matrices. Nevertheless for all couple $(i,j)
\in \{1, \ldots, n\}^2$, one can consider the composite $f_{ij} = 
\pr_{ij} \circ f$ where $\pr_{ij} : M_n(K) \to K$ that takes a matrix 
$M$ to its $(i,j)$-th entry. The maps $f_{ij}$'s are differentiable and 
their differentials generally are surjective (\emph{i.e.} they do not 
vanish). Let $M$ be a matrix known at some finite precision such that 
$f'_{ij} (M) \neq 0$ for all $(i,j)$. We can then apply the forward 
and/or the backward computation and get $f_{ij}(M)$ together with its 
precision. Putting this together, we get the whole matrix $f(M)$ 
together with a sharp precision datum on each entry.

The study of the this example actually suggests another solution to 
tackle the issue of non-surjectivity. Indeed, remark that our $f$ above 
had not a surjective differential simply because its codomain was too 
large: if we had replaced $f : (K^n)^n \to M_n(K)$ by $g : (K^n)^n \to 
S_n(K)$ (where $S_n(K)$ denotes the $K$-vector space of symmetric matrix 
over $K$ of size $n$) defined in the same way, our problem would 
disappear. Of course the image of a general $f$ is rarely a sub vector 
space of $F$ but it is often a sub-$K$-manifold of $F$. 
% Using the results of Appendix

We conclude this Section by remarking that the two strategies designed 
above (\emph{i.e.} forward and backward computations) share some 
similarities with usual floating point arithmetics over the reals. 
Indeed, roughly speaking, in each setting, we begin by choosing a large 
precision, we do all our computations up to this precision understood 
that when we are not sure about some digit, we choose it ``at random'' 
or using good heuristics. The main difference is that, in the 
ultrametric setting, we are able (under some mild hypothesis) to 
quantify the precision we need at each individual step in order to be 
sure that the final result is correct up to the required precision.
